If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization

### 2025-08-14 GroupPolarizationStatmod/paper/ms.tex
restructured Methods; ready for final drafting

Metascience frames our analysis since metascientific theory provides a framework for
our goal of calculating the false discovery rate across experimental conditions,
journal articles, and the ten-paper corpus. The false discovery rate is then
the Type I error rate relative to the rate of positive discoveries, true or false.
We calculate via simulation and data fitting
the Type I error rate for the experimental designs used in experimental conditions
that, under a preliminary large-$N$ model analysis, generated group
polarization upon ordinal measurement of simple consensus latent opinion 
distributions. These can then be aggregated to estimate the false discovery rate
for each journal article and for the ten-article corpus, which is representative of the
false discovery rate for group polarization studies.

\subsection{How study design can induce false detections of group polarization}
\label{subsec:false-detections}

An opinion is never measured directly, only a behavior that indicates an opinion
in a behavioral experiment or survey.  For example, the location of a mark
that a participant makes is measured, not the complex neural activity that
underlies forming opinions (Figure~\ref{fig:experiment_model}A) \mt{Some of this
needs to be Intro'd}.
Theoretically, an internal opinion in the form of neural activity is a
\textit{latent opinion}, from the Latin \emph{latentem} meaning ``lie
hidden''.  Participants report \text{observed opinions}, typically as one of
a finite number of ranked, successive integer values that form the
\emph{selection set}. The minimum possible value in the selection set might
mean ``strongly disagree'' and the maximum value represents ``strongly
agree''.

% \begin{wrapfigure}{r}{0.5\textwidth}
\begin{figure}
  \caption{\textbf{Opinion measurement model and spurious group polarization.} 
  Latent model participant opinions are drawn from a normal distribution with the
  listed parameters (A,C). The pre-deliberation latent distribution has high
  variance, but appears polarized when measured via integration (A).  The latent
  distribution's variance is expected to decrease during deliberation (B) via
  consensus, but in this example the latent mean is static, 
$\mupre = \mupost$. However, when the post-deliberation distribution is measured,
the simulated observed mean increases, i.e., spurious
group polarization occurs (C).}
  \centering
  \includegraphics[width=0.65\textwidth]{Figures/Model/latent-ordinal-distros.pdf}
  \label{fig:distros}
\end{figure}
% \end{wrapfigure}

Group polarization experiments share a common experimental design where,
first, participant opinions are gathered on a topic of interest. Next, groups
are formed that have some initial opinion extremism. Participants deliberate
or discuss the topic for some set time. After the set time, participants
again give their opinion. Group polarization is said to occur if the
mean group opinion became more extreme, i.e., if the magnitude of the mean
group opinion is greater \emph{post-deliberation} compared to
\emph{pre-deliberation}. For example,
\cite{Moscovici1969} asked Parisian high school students to indicate their
degree of agreement with the statement, ``American economic aid is always
used for political pressure'' (Figure~\ref{fig:experiment_model}A). There
were seven options in the selection set, -3 indicating ``strongly disagree'',
+3 indicating ``strongly agree'', and 0 was neutral. Moscovici and Zavalloni
report that the mean pre-deliberation opinion was -0.61 and the mean
post-deliberation opinion was -1.09 (Moscovici and Zavalloni, 1969; Table 4,
p. 132). This is a -0.48 shift to more extreme disagreement with the 
statement, reported by Moscovici and Zavalloni as statistically significant
in Table 5 on p. 132.

@@

>>>>

(METHODS OVERVIEW)

\subsection{Probability theory of spurious group polarization}

@@

Metascience provides a framework for
calculating the false discovery rate across experimental conditions,
aggregated to the level of journal articles and the ten-paper corpus. 
The false discovery rate is then the Type I error rate relative to the rate of
positive discoveries, true or false.
We calculate via simulation and data fitting
the Type I error rate for the experimental designs used in experimental conditions
that, under a preliminary large-$N$ model analysis, generated group
polarization upon ordinal measurement of simple consensus latent opinion 
distributions. These can then be aggregated to estimate the false discovery rate
for each journal article and for the ten-article corpus, which is representative of the
false discovery rate for group polarization studies.
